---
title: "R Programming Example: Create a dice-rolling program"
author: "Mary Amon"
date: "2023-03-15"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
set.seed(123) # for reproducibility
```


Adapted from code examples in [Hands-On Programming with R by Garrett Grolemund](https://rstudio-education.github.io/hopr/)

In this example, we will

* 1. Create a 6-sided die object
* 2. Create a function that will roll any number of 6-sided dice
* 3. Use the function within a for loop to roll the dice 100 times
* 4. Create a visualization of the 100 dice rolls

Note that this example only uses the Base R language without any imported packages.
In practice, you would import specialized packages to manage data wrangling and data visualization.

# 1. Create a 6-sided die

The : operator will create an atomic vector, a one-dimensional set of 1 data type.
The <- operator is the assignment operator to save data to an R object.

```{r}
# the operator : to create a 1-dimensional vector with values 1,2,3,4,5,6
# the operator <- assigns the vector to the R object called "die"
die <- 1:6
```

# 2. Create a function that will roll any number of 6 sided dice

Base R has a function called sample() that we can use to "roll" the die.
sample() takes a sample from a set of elements.

Every parameter in every R function has a name. You can specify which data should be assigned to which argument. This is optional, but aids readability and prevents errors as you use functions with many parameters. 

To roll 1 die, the parameters are:

* x: the elements to choose from (die)
* size: the number of items to choose (1 roll)

Note: by default, sample() will choose a value from x with equal probability for each element. 

```{r}
sample(x=die,size = 1)
```

If we want to roll 2 dice, we need to pass an additional parameter called replace.

When we pick a value from [1,2,3,4,5,6] for the first time, we want to put back that value so that when we pick the second value, all the values are still in the set. 

To roll 2 dice, the parameters are: 

* x: the elements to choose from (die)
* size: the number of items to choose (2 rolls)
* replace: put the value back when selecting the next element (TRUE)

```{r}
sample(die,size = 2, replace = TRUE)
```

We can put the previous bit of code within a function that will let us roll any number of 6-sided dice we want. 

Functions in R are another type of R object. Instead of containing data, they contain code.

Every function in R has 3 basic parts. Here are the parts for our roll function:

* 1. **Name:** use the <- operator to assign the function to the R object roll
* 2. **Body of code:** within curly braces, define the die and call sample() to "roll" dice
* 3. **Parameters:** within parenthesis, add parameter name numDice for how many dice we want to roll

```{r}
roll <- function(numDice){
  die <- 1:6
  sample(die, size=numDice, replace = TRUE)
}
```

Now, we can call our function to play any games that have dice!

```{r}
#Craps have 2 dice
crapsDice = 2
crapsRoll = roll(crapsDice)
print(crapsRoll)
```

```{r}
#Yahtzee has 5 dice
yahtzeeDice = 5
yahtzeeRoll = roll(yahtzeeDice)
print(yahtzeeRoll)
```

# 3. Use the function within a for loop to roll the dice 100 times

Like other programming languages, R has loops for repeating a task.

for loop in R resembles for loop in other languages, as it uses:

* the keyword "for"
* a parameter indicating how to loop
* a body of code to loop over. 

Similar to Python, the parameter for looping is iterating through a list. 

To save output our loop to roll dice, we need to save the result as it runs. We will use that in a dataframe, which is a 2-dimensional version of a list. A dataframe groups vectors together in a two-dimensional table. Each column of a data frame can be a different type of data, but within a column, every cell must be the same type. 

Note: If you use a for loop in R, you will often run into "Hey, you shouldn't be using a for loop!". That being said, sometimes you just need to use a loop, you know?

```{r}
data <- data.frame(die1=roll(1),die2=roll(1)) #initialize dataframe with 1 roll
for (i in 1:99){ #iterate through the values in the atomic vector 1 to 100
  data[nrow(data) + 1, ] <- roll(crapsDice) # roll the dice and store it in the next row
}
head(data,3) # inspect the table as a sanity check
```

# 4. Create a visualization of the 100 dice rolls

We can access columns in a dataframe with the $ character.
Let's add a new column called "total" adding up the value from die1 and die2

```{r}
data$total <- data$die1 + data$die2
head(data,3)
```

We can use the Base R function aggregate() to group all the 1s, 2s, etc.

The parameters of aggregate() are:

* x: the data to be split into groups by grouping
* by: the list of grouping elements
* FUN: a function to compute the summary statistics. Here we use length, which is "length" in the sense of the length of a list. You could also use FUN = mean, median, sum, etc. 

```{r}
totalcounts <-aggregate(x=data$total, by = list(data$total), FUN = length)
print(totalcounts)
```

We can graph this distribution using Base R barplot() function. 

Parameters for barplot():

* height: a vector of values describing the height of each bar in the barplot should be
* names.arg: a vactor of names to be plotted below each bar

```{r}
myplot <- barplot(height=totalcounts$x,names.arg=totalcounts$Group.1)
myplot
```


# Bonus: Compare our frequency table to the frequency of all possible combinations

Base R has a function called expand.grid that returns a dataframe of all possible combinations.

```{r}
rolls <- expand.grid(die1=die,die2=die)
head(rolls,3)
```

We can access columns in a dataframe with the $ character.
Let's add a new column called "total" adding up the value from die1 and die2

```{r}
rolls$total <- rolls$die1 + rolls$die2
head(rolls,3)
```

Base R aggregate() to group all the 1s, 2s, etc, in the total
```{r}
freq_df <-aggregate(rolls$total, by = list(rolls$total), FUN = length)
freq_df
```

Compare our simulated rolls above to the distribution of all possible combinations.

```{r}
library(ggplot2)
library(gridExtra)

totalcounts$Group.1 <- as.factor(totalcounts$Group.1)
myplot2 <- ggplot(data=totalcounts,aes(x=Group.1,y=x)) + 
  geom_bar(stat="identity") +
  xlab("Dice Total") +
  ylab("Frequency")

freq_df$Group.1 <- as.factor(freq_df$Group.1)
distplot <- ggplot(data=freq_df,aes(x=Group.1,y=x)) +
  geom_bar(stat="identity") +
  xlab("Dice Total") +
  ylab("Frequency")

grid.arrange(myplot2,distplot)
```

